---
title: "Adjusted sequential p-values"
author: "Yujie Zhao, Qi Liu, Linda Sun, Keaven Anderson"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
# bibliography: "example.bib"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Adjusted sequential p-values}
---

```{r, message = FALSE, warning = FALSE}
library(tibble)
library(gt)
library(gsDesign)
library(dplyr)
library(mvtnorm)
library(ggplot2)
library(wpgsd)
```

# Example overview

In a 2-arm controlled clinical trial example with one primary endpoint, there are 3 patient populations defined by the status of two biomarkers A and B: 

- biomarker A positive, 
- biomarker B positive, 
- overall population. 

The 3 primary elementary hypotheses are: 

- $H_1$: the experimental treatment is superior to the control in the biomarker A positive population;
- \item $H_2$: the experimental treatment is superior to the control in the biomarker B positive population;
- \item $H_3$: the experimental treatment is superior to the control in the overall population.

Assume an interim analysis and a final analysis are planned for the study and the number of events are listed as
```{r}
event_tbl <- tribble(
  ~population, ~analysis, ~event, 
  "A positive",     1,       80, 
  "B positive",     1,       88,
  "AB positive",    1,       64,
  "overall",        1,       180,
  "A positive",     2,       160,
  "B positive",     2,       176,
  "AB positive",    2,       128,
  "overall",        2,       360,
)
```

The observed p-values are
```{r}
obs_tbl <- tribble(
  ~hypothesis, ~analysis, ~obs_p,
    "H1",          1,       0.02,
    "H2",          1,       0.01,
    "H3",          1,       0.006,
    "H1",          2,       0.015,
    "H2",          2,       0.012,
    "H3",          2,       0.004) %>% 
  mutate(obs_Z = -qnorm(obs_p))

obs_tbl %>% 
  gt() %>% 
  tab_header(title = "Nominal p-values")

p_obs_IA <- (obs_tbl %>% filter(analysis == 1))$obs_p 
p_obs_FA <- (obs_tbl %>% filter(analysis == 2))$obs_p 
```


The information fraction of $H_1, H_2, H_3$ at IA is
```{r}
IF_IA <- c(
  ((event_tbl %>% filter(analysis == 1, population == "A positive"))$event + (event_tbl %>% filter(analysis == 1, population == "overall"))$event)/
    ((event_tbl %>% filter(analysis == 2, population == "A positive"))$event + (event_tbl %>% filter(analysis == 2, population == "overall"))$event),
  ((event_tbl %>% filter(analysis == 1, population == "B positive"))$event + (event_tbl %>% filter(analysis == 1, population == "overall"))$event)/
    ((event_tbl %>% filter(analysis == 2, population == "B positive"))$event + (event_tbl %>% filter(analysis == 2, population == "overall"))$event),
  ((event_tbl %>% filter(analysis == 1, population == "AB positive"))$event + (event_tbl %>% filter(analysis == 1, population == "overall"))$event)/
    ((event_tbl %>% filter(analysis == 2, population == "AB positive"))$event + (event_tbl %>% filter(analysis == 2, population == "overall"))$event)
)

IF_IA
```

we assign the initial weights of $H_1, H_2, H_3$ as$\left(w_1(I), w_2(I), w_3(I) \right) = (0.3, 0.3, 0.4).$ And its multiplicity strategy is visualized in below. If $H_1$ is rejected, then $3/7$ local significance level $\alpha_1$ will be propagated to $H_2$, and $4/7$ will go to $H_3$. If $H_3$ is rejected, then half of $\alpha_3$ goes to $H_1$, and half goes to $H_2$. 
```{r}
# Transition matrix in Figure A1
m <- matrix(c(0,   3/7, 4/7,
              3/7, 0,   4/7,
              1/2, 1/2, 0),nrow = 3, byrow = TRUE)
# Initial weights
w <- c(0.3, 0.3, 0.4)
```

```{r}
name_hypotheses <- c("H1: Biomarker A positive", "H2: Biomarker B positive",  "H3: Overall Population")

hplot <- gMCPLite::hGraph(
  3,alphaHypotheses=w,m=m,
  nameHypotheses=name_hypotheses, trhw=.2, trhh=.1, 
  digits=5, trdigits=3, size=5, halfWid=1, halfHgt=0.5,      
  offset=0.2 , trprop= 0.4,  
  fill=as.factor(c(2,3,1)),
  palette = c("#BDBDBD", "#E0E0E0", "#EEEEEE"),
  wchar = "w") 
hplot
```

The correlation of the 6 statistisc (2 analysis $\times$ 3 hypothesis) are
```{r}
# Event count of intersection of paired hypotheses - Table 2
# H1, H2: Hypotheses intersected.
# (1, 1) represents counts for hypothesis 1
# (1, 2) for counts for the intersection of hypotheses 1 and 2
event <- tribble(
  ~H1, ~H2, ~Analysis, ~Event,
  1,    1,      1,     event_tbl %>% filter(analysis == 1, population == "A positive") %>% select(event) %>% as.numeric(),
  2,    2,      1,     event_tbl %>% filter(analysis == 1, population == "B positive") %>% select(event) %>% as.numeric(),
  3,    3,      1,     event_tbl %>% filter(analysis == 1, population == "overall") %>% select(event) %>% as.numeric(),
  1,    2,      1,     event_tbl %>% filter(analysis == 1, population == "AB positive") %>% select(event) %>% as.numeric(),
  1,    3,      1,     event_tbl %>% filter(analysis == 1, population == "A positive") %>% select(event) %>% as.numeric(),
  2,    3,      1,     event_tbl %>% filter(analysis == 1, population == "B positive") %>% select(event) %>% as.numeric(),
  1,    1,      2,     event_tbl %>% filter(analysis == 2, population == "A positive") %>% select(event) %>% as.numeric(),
  2,    2,      2,     event_tbl %>% filter(analysis == 2, population == "B positive") %>% select(event) %>% as.numeric(),
  3,    3,      2,     event_tbl %>% filter(analysis == 2, population == "overall") %>% select(event) %>% as.numeric(),
  1,    2,      2,     event_tbl %>% filter(analysis == 2, population == "AB positive") %>% select(event) %>% as.numeric(),
  1,    3,      2,     event_tbl %>% filter(analysis == 2, population == "A positive") %>% select(event) %>% as.numeric(),
  2,    3,      2,     event_tbl %>% filter(analysis == 2, population == "B positive") %>% select(event) %>% as.numeric()) 
event  

# Generate correlation from events
gs_corr <- wpgsd::generate_corr(event)
gs_corr %>% round(2)
```

# Sequantial p-value

## IA

```{r, wpgsdSeqIA, message=FALSE}
seq_p_IA_H123 <- calc_seq_p(
  test_analysis = 1,            # stage of interest
  test_hypothesis = "H1, H2, H3",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 2,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = gsDesign::sfHSD,
  spending_fun_par = -4,
  info_frac = c(min(IF_IA), 1),
  interval = c(1e-4, 0.15)
  )

seq_p_IA_H12 <- calc_seq_p(
  test_analysis = 1,            # stage of interest
  test_hypothesis = "H1, H2",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 2,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = gsDesign::sfHSD,
  spending_fun_par = -4,
  info_frac = c(min(IF_IA), 1),
  interval = c(1e-4, 0.15)
  )

seq_p_IA_H13 <- calc_seq_p(
  test_analysis = 1,            # stage of interest
  test_hypothesis = "H1, H3",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 2,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = gsDesign::sfHSD,
  spending_fun_par = -4,
  info_frac = c(min(IF_IA), 1),
  interval = c(1e-4, 0.15)
  )

seq_p_IA_H23 <- calc_seq_p(
  test_analysis = 1,            # stage of interest
  test_hypothesis = "H2, H3",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 2,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = gsDesign::sfHSD,
  spending_fun_par = -4,
  info_frac = c(min(IF_IA), 1),
  interval = c(1e-4, 0.15)
  )

seq_p_IA_H1 <- calc_seq_p(
  test_analysis = 1,            # stage of interest
  test_hypothesis = "H1",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 2,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = gsDesign::sfHSD,
  spending_fun_par = -4,
  info_frac = c(min(IF_IA), 1),
  interval = c(1e-4, 0.2)
  )

seq_p_IA_H2 <- calc_seq_p(
  test_analysis = 1,            # stage of interest
  test_hypothesis = "H2",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 2,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = gsDesign::sfHSD,
  spending_fun_par = -4,
  info_frac = c(min(IF_IA), 1),
  interval = c(1e-4, 0.2)
  )

seq_p_IA_H3 <- calc_seq_p(
  test_analysis = 1,            # stage of interest
  test_hypothesis = "H3",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 2,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = gsDesign::sfHSD,
  spending_fun_par = -4,
  info_frac = c(min(IF_IA), 1),
  interval = c(1e-4, 0.2)
  )
```

```{r, weightedBonferroniSeqIA, message=FALSE}
seq_p_IA_H123_B <- calc_seq_p(
  test_analysis = 1,            # stage of interest
  test_hypothesis = "H1, H2, H3",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 0,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = list(gsDesign::sfHSD, gsDesign::sfHSD, gsDesign::sfHSD),
  spending_fun_par = list(-4, -4, -4),
  info_frac = list(c(IF_IA[1], 1), c(IF_IA[2], 1), c(IF_IA[2], 1)),
  interval = c(1e-4, 0.2)
  )

seq_p_IA_H12_B <- calc_seq_p(
  test_analysis = 1,            # stage of interest
  test_hypothesis = "H1, H2",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 0,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = list(gsDesign::sfHSD, gsDesign::sfHSD, gsDesign::sfHSD),
  spending_fun_par = list(-4, -4, -4),
  info_frac = list(c(IF_IA[1], 1), c(IF_IA[2], 1), c(IF_IA[2], 1)),
  interval = c(1e-4, 0.2)
  )

seq_p_IA_H13_B <- calc_seq_p(
  test_analysis = 1,            # stage of interest
  test_hypothesis = "H1, H3",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 0,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = list(gsDesign::sfHSD, gsDesign::sfHSD, gsDesign::sfHSD),
  spending_fun_par = list(-4, -4, -4),
  info_frac = list(c(IF_IA[1], 1), c(IF_IA[2], 1), c(IF_IA[2], 1)),
  interval = c(1e-4, 0.2)
  )

seq_p_IA_H23_B <- calc_seq_p(
  test_analysis = 1,            # stage of interest
  test_hypothesis = "H2, H3",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 0,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = list(gsDesign::sfHSD, gsDesign::sfHSD, gsDesign::sfHSD),
  spending_fun_par = list(-4, -4, -4),
  info_frac = list(c(IF_IA[1], 1), c(IF_IA[2], 1), c(IF_IA[2], 1)),
  interval = c(1e-4, 0.2)
  )

seq_p_IA_H1_B <- calc_seq_p(
  test_analysis = 1,            # stage of interest
  test_hypothesis = "H1",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 0,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = list(gsDesign::sfHSD, gsDesign::sfHSD, gsDesign::sfHSD),
  spending_fun_par = list(-4, -4, -4),
  info_frac = list(c(IF_IA[1], 1), c(IF_IA[2], 1), c(IF_IA[2], 1)),
  interval = c(1e-4, 0.2)
  )

seq_p_IA_H2_B <- calc_seq_p(
  test_analysis = 1,            # stage of interest
  test_hypothesis = "H2",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 0,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = list(gsDesign::sfHSD, gsDesign::sfHSD, gsDesign::sfHSD),
  spending_fun_par = list(-4, -4, -4),
  info_frac = list(c(IF_IA[1], 1), c(IF_IA[2], 1), c(IF_IA[2], 1)),
  interval = c(1e-4, 0.2)
  )

seq_p_IA_H3_B <- calc_seq_p(
  test_analysis = 1,            # stage of interest
  test_hypothesis = "H3",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 0,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = list(gsDesign::sfHSD, gsDesign::sfHSD, gsDesign::sfHSD),
  spending_fun_par = list(-4, -4, -4),
  info_frac = list(c(IF_IA[1], 1), c(IF_IA[2], 1), c(IF_IA[2], 1)),
  interval = c(1e-4, 0.2)
  )
```

## FA

```{r, wpgsdSeqFA, message=FALSE}
seq_p_FA_H123 <- calc_seq_p(
  test_analysis = 2,            # stage of interest
  test_hypothesis = "H1, H2, H3",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 2,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = gsDesign::sfHSD,
  spending_fun_par = -4,
  info_frac = c(min(IF_IA), 1),
  interval = c(1e-4, 0.15)
  )

seq_p_FA_H12 <- calc_seq_p(
  test_analysis = 2,            # stage of interest
  test_hypothesis = "H1, H2",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 2,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = gsDesign::sfHSD,
  spending_fun_par = -4,
  info_frac = c(min(IF_IA), 1),
  interval = c(1e-4, 0.15)
  )

seq_p_FA_H13 <- calc_seq_p(
  test_analysis = 2,            # stage of interest
  test_hypothesis = "H1, H3",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 2,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = gsDesign::sfHSD,
  spending_fun_par = -4,
  info_frac = c(min(IF_IA), 1),
  interval = c(1e-4, 0.15)
  )

seq_p_FA_H23 <- calc_seq_p(
  test_analysis = 2,            # stage of interest
  test_hypothesis = "H2, H3",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 2,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = gsDesign::sfHSD,
  spending_fun_par = -4,
  info_frac = c(min(IF_IA), 1),
  interval = c(1e-4, 0.15)
  )

seq_p_FA_H1 <- calc_seq_p(
  test_analysis = 2,            # stage of interest
  test_hypothesis = "H1",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 2,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = gsDesign::sfHSD,
  spending_fun_par = -4,
  info_frac = c(min(IF_IA), 1),
  interval = c(1e-4, 0.2)
  )

seq_p_FA_H2 <- calc_seq_p(
  test_analysis = 2,            # stage of interest
  test_hypothesis = "H2",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 2,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = gsDesign::sfHSD,
  spending_fun_par = -4,
  info_frac = c(min(IF_IA), 1),
  interval = c(1e-4, 0.2)
  )

seq_p_FA_H3 <- calc_seq_p(
  test_analysis = 2,            # stage of interest
  test_hypothesis = "H3",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 2,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = gsDesign::sfHSD,
  spending_fun_par = -4,
  info_frac = c(min(IF_IA), 1),
  interval = c(1e-4, 0.2)
  )
```

```{r, weightedBonferroniSeqFA, message=FALSE}
seq_p_FA_H123_B <- calc_seq_p(
  test_analysis = 2,            # stage of interest
  test_hypothesis = "H1, H2, H3",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 0,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = list(gsDesign::sfHSD, gsDesign::sfHSD, gsDesign::sfHSD),
  spending_fun_par = list(-4, -4, -4),
  info_frac = list(c(IF_IA[1], 1), c(IF_IA[2], 1), c(IF_IA[2], 1)),
  interval = c(1e-4, 0.2)
  )

seq_p_FA_H12_B <- calc_seq_p(
  test_analysis = 2,            # stage of interest
  test_hypothesis = "H1, H2",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 0,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = list(gsDesign::sfHSD, gsDesign::sfHSD, gsDesign::sfHSD),
  spending_fun_par = list(-4, -4, -4),
  info_frac = list(c(IF_IA[1], 1), c(IF_IA[2], 1), c(IF_IA[2], 1)),
  interval = c(1e-4, 0.2)
  )

seq_p_FA_H13_B <- calc_seq_p(
  test_analysis = 2,            # stage of interest
  test_hypothesis = "H1, H3",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 0,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = list(gsDesign::sfHSD, gsDesign::sfHSD, gsDesign::sfHSD),
  spending_fun_par = list(-4, -4, -4),
  info_frac = list(c(IF_IA[1], 1), c(IF_IA[2], 1), c(IF_IA[2], 1)),
  interval = c(1e-4, 0.2)
  )

seq_p_FA_H23_B <- calc_seq_p(
  test_analysis = 2,            # stage of interest
  test_hypothesis = "H2, H3",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 0,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = list(gsDesign::sfHSD, gsDesign::sfHSD, gsDesign::sfHSD),
  spending_fun_par = list(-4, -4, -4),
  info_frac = list(c(IF_IA[1], 1), c(IF_IA[2], 1), c(IF_IA[2], 1)),
  interval = c(1e-4, 0.2)
  )

seq_p_FA_H1_B <- calc_seq_p(
  test_analysis = 2,            # stage of interest
  test_hypothesis = "H1",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 0,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = list(gsDesign::sfHSD, gsDesign::sfHSD, gsDesign::sfHSD),
  spending_fun_par = list(-4, -4, -4),
  info_frac = list(c(IF_IA[1], 1), c(IF_IA[2], 1), c(IF_IA[2], 1)),
  interval = c(1e-4, 0.2)
  )

seq_p_FA_H2_B <- calc_seq_p(
  test_analysis = 2,            # stage of interest
  test_hypothesis = "H2",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 0,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = list(gsDesign::sfHSD, gsDesign::sfHSD, gsDesign::sfHSD),
  spending_fun_par = list(-4, -4, -4),
  info_frac = list(c(IF_IA[1], 1), c(IF_IA[2], 1), c(IF_IA[2], 1)),
  interval = c(1e-4, 0.2)
  )

seq_p_FA_H3_B <- calc_seq_p(
  test_analysis = 2,            # stage of interest
  test_hypothesis = "H3",
  p_obs = p_obs_IA,             # observed p-value
  alpha_spending_type = 0,
  n_analysis = 2,
  initial_weight = w,
  transition_mat = m,
  z_corr = gs_corr,
  spending_fun = list(gsDesign::sfHSD, gsDesign::sfHSD, gsDesign::sfHSD),
  spending_fun_par = list(-4, -4, -4),
  info_frac = list(c(IF_IA[1], 1), c(IF_IA[2], 1), c(IF_IA[2], 1)),
  interval = c(1e-4, 0.2)
  )
```


# Adjusted-Sequential p-value

## IA

```{r, wpgsdAdjSeqIA}
adj_seq_IA_H1 <- max(seq_p_IA_H123, seq_p_IA_H12, seq_p_IA_H13, seq_p_IA_H1)
adj_seq_IA_H2 <- max(seq_p_IA_H123, seq_p_IA_H12, seq_p_IA_H23, seq_p_IA_H2)
adj_seq_IA_H3 <- max(seq_p_IA_H123, seq_p_IA_H13, seq_p_IA_H23, seq_p_IA_H3)

cat("The adjusted-sequential p-value of H1, H2, H3 in IA via WPGSD is", adj_seq_IA_H1, adj_seq_IA_H2, adj_seq_IA_H3, "\n")
```

```{r, weightedBonferroniAdjSeqIA}
adj_seq_IA_H1_B <- max(seq_p_IA_H123_B, seq_p_IA_H12_B, seq_p_IA_H13_B, seq_p_IA_H1_B)
adj_seq_IA_H2_B <- max(seq_p_IA_H123_B, seq_p_IA_H12_B, seq_p_IA_H23_B, seq_p_IA_H2_B)
adj_seq_IA_H3_B <- max(seq_p_IA_H123_B, seq_p_IA_H13_B, seq_p_IA_H23_B, seq_p_IA_H3_B)

cat("The adjusted-sequential p-value of H1, H2, H3 in FA via weighted Bonferroni is", adj_seq_IA_H1_B, adj_seq_IA_H2_B, adj_seq_IA_H3_B, "\n")
```


## FA 

### WPGSD
```{r, wpgsdAdjSeqFA}
adj_seq_FA_H1 <- max(seq_p_FA_H123, seq_p_FA_H12, seq_p_FA_H13, seq_p_FA_H1)
adj_seq_FA_H2 <- max(seq_p_FA_H123, seq_p_FA_H12, seq_p_FA_H23, seq_p_FA_H2)
adj_seq_FA_H3 <- max(seq_p_FA_H123, seq_p_FA_H13, seq_p_FA_H23, seq_p_FA_H3)

cat("The adjusted-sequential p-value of H1, H2, H3 in FA via WPGSD is", adj_seq_FA_H1, adj_seq_FA_H2, adj_seq_FA_H3, "\n")
```

```{r, weightedBonferroniAdjSeqFA}
adj_seq_FA_H1_B <- max(seq_p_FA_H123_B, seq_p_FA_H12_B, seq_p_FA_H13_B, seq_p_FA_H1_B)
adj_seq_FA_H2_B <- max(seq_p_FA_H123_B, seq_p_FA_H12_B, seq_p_FA_H23_B, seq_p_FA_H2_B)
adj_seq_FA_H3_B <- max(seq_p_FA_H123_B, seq_p_FA_H13_B, seq_p_FA_H23_B, seq_p_FA_H3_B)

cat("The adjusted-sequential p-value of H1, H2, H3 in FA via weighted Bonferroni is", adj_seq_FA_H1_B, adj_seq_FA_H2_B, adj_seq_FA_H3_B, "\n")
```

# Summary

```{r}
ans <- tribble(
  ~Analysis, ~Hypothesis, ~`Sequantial p-values of WPGSD`,  ~`Sequantial p-values of Weighted Bonferroni`, ~`Adjusted-sequantial p-values of WPGSD`, ~`Adjusted-sequantial p-values of Weighted Bonferroni`, 
  "IA",       "H123",       seq_p_IA_H123,      seq_p_IA_H123_B,    NA,                    NA,                                
  "IA",       "H12",        seq_p_IA_H12,       seq_p_IA_H12_B,     NA,                    NA,                                     
  "IA",       "H13",        seq_p_IA_H13,       seq_p_IA_H13_B,     NA,                    NA,                                  
  "IA",       "H23",        seq_p_IA_H23,       seq_p_IA_H23_B,     NA,                    NA,                                       
  "IA",       "H1",         seq_p_IA_H1,        seq_p_IA_H1_B,      adj_seq_IA_H1,         adj_seq_IA_H1_B,                     
  "IA",       "H2",         seq_p_IA_H2,        seq_p_IA_H2_B,      adj_seq_IA_H2,         adj_seq_IA_H2_B,                     
  "IA",       "H3",         seq_p_IA_H3,        seq_p_IA_H3_B,      adj_seq_IA_H3,         adj_seq_IA_H3_B,                     
  "FA",       "H123",       seq_p_FA_H123,      seq_p_FA_H123_B,    NA,                    NA,                                  
  "FA",       "H12",        seq_p_FA_H12,       seq_p_FA_H12_B,     NA,                    NA,                                  
  "FA",       "H13",        seq_p_FA_H13,       seq_p_FA_H13_B,     NA,                    NA,                                  
  "FA",       "H23",        seq_p_FA_H23,       seq_p_FA_H23_B,     NA,                    NA,                                  
  "FA",       "H1",         seq_p_FA_H1,        seq_p_FA_H1_B,      adj_seq_FA_H1,         adj_seq_FA_H1_B,                     
  "FA",       "H2",         seq_p_FA_H2,        seq_p_FA_H2_B,      adj_seq_FA_H2,         adj_seq_FA_H2_B,                     
  "FA",       "H3",         seq_p_FA_H3,        seq_p_FA_H3_B,      adj_seq_FA_H3,         adj_seq_FA_H3_B
  ) 

ans %>% 
  gt::gt() %>% 
  gt::tab_style_body(
    columns = where(is.numeric),
    style = cell_fill(color = "pink"),
    fn = function(x) x <= 0.025
  ) %>% 
  gt::fmt_number(columns = 3:6, decimals = 4) %>% 
  gt::tab_header(title = "(Adjusted-) sequantial p-values",
                 subtitle = "Multiple experiment arm vs. common control") # %>% gt::as_latex()
```


