---
title: "Correlated test statistics"
author: "Chenguang Zhang, Yujie Zhao and Keaven Anderson"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
    code_fold: hide
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Correlated test statistics}
bibliography: wpgsd.bib
---

The weighted parametric group sequential design (WPGSD) (@anderson2022unified) approach allows one to take advantage of the known correlation structure 
in constructing efficacy bounds to control family-wise error rate (FWER) for a group sequential design. 
Here correlation may be due to common observations in nested populations, due to common observations in overlapping populations, 
or due to common observations in the control arm. 

# Methodologies to calculate correlations

Suppose that in a group sequential trial there are $m$ elementary null hypotheses $H_i$, $i \in I={1,...,m},$ and there are $K$ analyses. 
Let $k=1,2,...K$ be the index for the interim analyses and final analyses. 
For any nonempty set $J \subseteq I$, we denote the intersection hypothesis $H_J=\cap_{j \in J}H_j$. We note that $H_I$ is the global null hypothesis.

We assume the plan is for all hypotheses to be tested at each of the $k$ planned analyses if the trial continues to the end for all hypotheses. We further assume that the distribution of the $m \times K$ tests of $m$ individual hypotheses at all $k$ analyses is multivariate normal with a completely known correlation matrix. 

Let $Z_{ik}$ be the standardized normal test statistic for hypothesis $i \in I$, analysis $1 \le k \le K$. 
Let $n_{ik}$ be the number of events collected cumulatively through stage $k$ for hypothesis $i$. 
Then $n_{i \wedge i',k \wedge k'}$ is the number of events included in both $Z_{ik}$ and $i$, $i' \in I$, $1 \le k$, $k' \le K$. 
The key for the parametric tests is to utilize the correlation among the test statistics. 
The correlation between $Z_{ik}$ and $Z_{i'k'}$ is
$$Corr(Z_{ik},Z_{i'k'})=\frac{n_{i \wedge i',k \wedge k'}}{\sqrt{n_{ik}\times n_{i'k'}}}$$. 

# Examples

We borrow an Example 1 from @anderson2022unified here.
Assume a two-arm controlled clinical trial with one primary endpoint and three patient populations defined by the status of two biomarkers, A and B:

* Biomarker A positive (population 1),
* Biomarker B positive (population 2),
* Overall population (population 3).

The 3 primary elementary null hypotheses are:

* **H1**: the experimental treatment is no different than control in population 1
* **H2**: the experimental treatment is no different than control in population 2
* **H3**: the experimental treatment is no different than control in the overall population
  
Assume an interim analysis and a final analysis are planned for the study. 
The number of events observed at each analysis in each population as well as the 
intersections of populations 1 and 2 to are assumed to be
```{r,message=FALSE}
library(dplyr)
library(tibble)
library(gt)
```

```{r}
event_tb <- tribble(
  ~Population, ~"Number of Event at IA", ~"Number of Event at FA",
  "Population 1", 100, 200,
  "Population 2", 110, 220,
  "Overlap of Population 1 and 2", 80, 160,
  "Overall Population", 225, 450
)
event_tb %>%
  gt() %>%
  tab_header(title = "Number of events in each population")
```

##  Correlation of tests for different populations at the same analysis

Let's consider the situation at the interim analysis.
The correlation between tests of H1 and H2 will be
$$Corr(Z_{11},Z_{21})=\frac{n_{1 \wedge 2,1 \wedge 1}}{\sqrt{n_{11}\times n_{21}}}=\frac{80}{\sqrt{100\times 110}}=0.76.$$

The 80 in the numerator is the overlap number of events of population 1 and population 2 at the interim analysis.

## Correlation for tests of the same population at different analyses

We consider the correlation between tests for population 1 at the IA and FA.
The correlation will thus be: 
$$Corr(Z_{11},Z_{12})=\frac{n_{1 \wedge 1,1 \wedge 2}}{\sqrt{n_{11}\times n_{12}}}=100/\sqrt{100\times 200}=0.71.$$

## Correlation of tests at different analyses for different populations

Next we consider the correlation between the tests of population 1 at the interim analysis and population 2 at the final analyses. 
The correlation will be
$$\text{Corr}(Z_{11},Z_{22})=\frac{n_{1 \wedge 1,2 \wedge 2}}{\sqrt{n_{11}\times n_{22}}}=\frac{80}{\sqrt{100\times 220}}=0.54.$$
The 80 in the numerator is the overlap number of events of population 1 in interim analysis and population 2 in final analysis;
i.e., the events in both populations 1 and 2 at the interim analysis.

# Generate the correlation matrix by `s7_generate_corr()`    

Now we know how to calculate the correlation values under different situations, and the `s7_generate_corr()` function was built based on this logic. 
We can directly calculate the complete correlation matrix for all tests via the function.

First, we need an event table including the event counts for testing individual hypotheses and pairwise intersection hypotheses for the study.

- `H1, H2` are each the index of a hypothesis. We will have the index for `H1` be $<=$ the index for `H2` to avoid duplication of pairs.
- `Analysis` indicates which of sequentially indexed analyses, interim and final, is being referred to.
- `Event` is the common number of events observed for testing both `H1` and `H2` at the specified `Analysis`.

For example: `H1=1`, `H2=1`, `Analysis=1`, `Event=100 `indicates that in the first population, there are 100 events for comparing the experimental treatment to control 
at the interim analysis. Thus, when `H1=H2`, we are looking at the number of events used for testing the simple hypothesis specified in `H1=H2`.

For testing an intersection hypothesis with `H1 != H2` and a given `Analysis` value, we are looking at the number of events that are used in testing both `H1` and `H2`.
For our example above, if `H1=1`, `H2=2`, `Analysis=2`, `Event=160` indicates that the number of overlapping 
cases for comparing the experimental treatment to the control in population 1 and 2 which is 160.

The column names needed for input to `s7_generate_corr()` are `H1`, `H2`, `Analysis`, `Event`.
We take the event counts from above accordingly.

```{r, message=FALSE}
library(wpgsd)
# The event table
event <- tibble::tribble(
  ~H1, ~H2, ~Analysis, ~Event,
  1, 1, 1, 100,
  2, 2, 1, 110,
  3, 3, 1, 225,
  1, 2, 1, 80,
  1, 3, 1, 100,
  2, 3, 1, 110,
  1, 1, 2, 200,
  2, 2, 2, 220,
  3, 3, 2, 450,
  1, 2, 2, 160,
  1, 3, 2, 200,
  2, 3, 2, 220
)
```

By specifying this as the S7 type `EventTable` we get input checks to ensure no missing values or inconsistencies.

```{r}
event <- as_event_table(event)
event
```

For human readability of the data table:

```{r}
event@data
```

Now we input the above event table to the function of `generate_corr_s7()`, 
and get the correlation matrix as follow.

```{r}
corr_mat <- generate_corr_s7(event)
corr_mat
```

For human readability, we can format the correlation matrix as a table.

```{r}
corr_mat@matrix
```

# References

