---
title: "Dunnett-type dose-finding with group sequential design"
author: "Keaven Anderson"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
bibliography: wpgsd.bib
vignette: >
  %\VignetteIndexEntry{Dunnett-type dose-finding with group sequential design}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  fig.align = "center"
)
```

```{r, message=FALSE, warning=FALSE}
library(wpgsd)
library(gsDesign)
library(gMCPLite)
library(dplyr)
library(tibble)
library(gt)
```

# Background

A common clinical trial design compares multiple experimental dose
groups against a shared control arm. This is known as a Dunnett-type
design (@dunnett1955multiple). When combined with a group sequential
design allowing an interim analysis for futility, the correlation
between test statistics for different dose-versus-control comparisons
arises naturally from the shared control arm.

The weighted parametric group sequential design (WPGSD) approach
(@anderson2022unified) exploits these known correlations to produce
tighter efficacy bounds compared to the Bonferroni approach. This
vignette demonstrates the WPGSD methodology for a trial with 4
experimental dose groups versus a common control, using overall survival
(OS) as the primary endpoint, with a group sequential design having an
interim analysis at approximately 40% information fraction.

# Trial Setup

## Hypotheses

We consider 4 hypotheses, each comparing an experimental dose group to
the common control:

-   $H_1$: Dose 1 vs. Control
-   $H_2$: Dose 2 vs. Control
-   $H_3$: Dose 3 vs. Control
-   $H_4$: Dose 4 vs. Control

## Multiplicity Strategy

We use equal initial weights ($w_i = 1/4$ for each hypothesis) and
equal transitions ($m_{ij} = 1/3$ for $i \neq j$). This symmetric
strategy reflects no prior preference among doses.

```{r, out.width="90%"}
# Transition matrix: equal transitions among 4 hypotheses
m <- matrix(1 / 3, nrow = 4, ncol = 4)
diag(m) <- 0

# Initial weights: equal allocation
w <- rep(1 / 4, 4)

# Multiplicity graph
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73")

nameHypotheses <- c(
  "H1: Dose 1 vs Control",
  "H2: Dose 2 vs Control",
  "H3: Dose 3 vs Control",
  "H4: Dose 4 vs Control"
)

hplot <- hGraph(4,
  alphaHypotheses = w,
  m = m,
  nameHypotheses = nameHypotheses,
  trhw = .2, trhh = .1,
  digits = 4, trdigits = 4, size = 5, halfWid = 1.2,
  halfHgt = 0.5, offset = 0.2, trprop = 0.35,
  fill = as.factor(1:4),
  palette = cbPalette,
  wchar = "w"
)
hplot
```

## Event Counts

We assume approximately balanced event counts across treatment arms with
an interim analysis (IA) at roughly 40% of the planned final analysis
(FA) events. For OS, events are deaths observed in each
dose-versus-control comparison.

| Treatment Arm  | Events at IA | Events at FA |
|:--------------:|:------------:|:------------:|
| Dose 1         | 24           | 60           |
| Dose 2         | 26           | 65           |
| Dose 3         | 25           | 63           |
| Dose 4         | 27           | 67           |
| Control        | 25           | 63           |

Table: Number of events at each analysis for each treatment arm.
IA: interim analysis (~40% information). FA: final analysis.

The diagonal entries of the event table represent total events for each
hypothesis (experimental arm + control), while off-diagonal entries
represent events shared between two comparisons (control arm events
only).

```{r}
# Event counts per arm
# IA: Dose 1=24, Dose 2=26, Dose 3=25, Dose 4=27, Control=25
# FA: Dose 1=60, Dose 2=65, Dose 3=63, Dose 4=67, Control=63

# Event table: diagonal = experimental + control; off-diagonal = control only
event <- tribble(
  ~H1, ~H2, ~Analysis, ~Event,
  # IA diagonal (total events per hypothesis = dose + control)
  1, 1, 1, 49,   # Dose 1 (24) + Control (25)
  2, 2, 1, 51,   # Dose 2 (26) + Control (25)
  3, 3, 1, 50,   # Dose 3 (25) + Control (25)
  4, 4, 1, 52,   # Dose 4 (27) + Control (25)
  # IA off-diagonal (shared events = control only)
  1, 2, 1, 25,
  1, 3, 1, 25,
  1, 4, 1, 25,
  2, 3, 1, 25,
  2, 4, 1, 25,
  3, 4, 1, 25,
  # FA diagonal (total events per hypothesis = dose + control)
  1, 1, 2, 123,  # Dose 1 (60) + Control (63)
  2, 2, 2, 128,  # Dose 2 (65) + Control (63)
  3, 3, 2, 126,  # Dose 3 (63) + Control (63)
  4, 4, 2, 130,  # Dose 4 (67) + Control (63)
  # FA off-diagonal (shared events = control only)
  1, 2, 2, 63,
  1, 3, 2, 63,
  1, 4, 2, 63,
  2, 3, 2, 63,
  2, 4, 2, 63,
  3, 4, 2, 63
)

event %>%
  gt() %>%
  tab_header(title = "Event Count Table")
```

# Correlation Structure

## Computing Correlations

The correlation between test statistics follows the formula
(@anderson2022unified):

$$\text{Corr}(Z_{ik}, Z_{i'k'}) = \frac{n_{i \wedge i', k \wedge k'}}{\sqrt{n_{ik} \cdot n_{i'k'}}}$$

where $n_{i \wedge i', k \wedge k'}$ denotes the number of events
included in both $Z_{ik}$ and $Z_{i'k'}$.

For the Dunnett-type design, the shared events between two different
dose-versus-control comparisons at the same analysis are the control arm
events. For example, at the interim analysis:

$$\text{Corr}(Z_{1,1}, Z_{2,1}) = \frac{n_{\text{control, IA}}}{\sqrt{n_{1,1} \cdot n_{2,1}}} = \frac{25}{\sqrt{49 \cdot 51}} \approx 0.500$$

```{r}
# Generate correlation matrix
corr <- generate_corr(event)

corr %>%
  as_tibble() %>%
  gt() %>%
  fmt_number(columns = everything(), decimals = 3) %>%
  tab_header(title = "Correlation Matrix (8 x 8)")
```

## Verification of Correlation Entries

We manually verify key entries of the correlation matrix to confirm
correctness for the 4-hypothesis Dunnett scenario.

```{r}
# --- Verification of key correlation entries ---

# Case 1: Same hypothesis, same analysis -> should be 1.0
cat("Corr(Z_1,1, Z_1,1) =", corr[1, 1], " (expected: 1.0)\n")

# Case 2: Different hypotheses, same analysis (IA)
# Corr(Z_1,1, Z_2,1) = n_control_IA / sqrt(n_H1_IA * n_H2_IA) = 25 / sqrt(49 * 51)
expected_12_ia <- 25 / sqrt(49 * 51)
cat("Corr(Z_1,1, Z_2,1) =", round(corr[1, 2], 6),
    " (expected:", round(expected_12_ia, 6), ")\n")

# Case 3: Different hypotheses, same analysis (FA)
# Corr(Z_1,2, Z_3,2) = n_control_FA / sqrt(n_H1_FA * n_H3_FA) = 63 / sqrt(123 * 126)
expected_13_fa <- 63 / sqrt(123 * 126)
cat("Corr(Z_1,2, Z_3,2) =", round(corr[5, 7], 6),
    " (expected:", round(expected_13_fa, 6), ")\n")

# Case 4: Same hypothesis, different analyses
# Corr(Z_1,1, Z_1,2) = n_H1_IA / sqrt(n_H1_IA * n_H1_FA) = 49 / sqrt(49 * 123)
expected_11_cross <- 49 / sqrt(49 * 123)
cat("Corr(Z_1,1, Z_1,2) =", round(corr[1, 5], 6),
    " (expected:", round(expected_11_cross, 6), ")\n")

# Case 5: Different hypotheses, different analyses
# Corr(Z_2,1, Z_4,2) = n_control_min(IA) / sqrt(n_H2_IA * n_H4_FA) = 25 / sqrt(51 * 130)
expected_24_cross <- 25 / sqrt(51 * 130)
cat("Corr(Z_2,1, Z_4,2) =", round(corr[2, 8], 6),
    " (expected:", round(expected_24_cross, 6), ")\n")
```

The correlation matrix for this 4-hypothesis Dunnett scenario has the
following structure:

-   **Same analysis, different hypotheses**: Correlation $\approx 0.50$
    at IA and $\approx 0.50$ at FA, driven by the shared control arm.
    With balanced arms, this approaches $n_C / (n_C + n_E) \approx 0.5$
    when $n_C \approx n_E$.
-   **Same hypothesis, different analyses**: Correlation $\approx 0.63$,
    the standard group sequential information fraction correlation
    $\sqrt{n_{ik}/n_{ik'}}$.
-   **Different hypotheses, different analyses**: Correlation
    $\approx 0.31$, combining both sources of correlation.

# Futility Bound

We use `gsDesign` to compute a common non-binding futility bound at the
interim analysis. This bound applies equally to all 4 dose arms. The
futility bound is based on a beta-spending approach using a
Hwang-Shih-DeCani spending function.

```{r}
# Group sequential design with non-binding futility
# Approximately 40% information fraction
# Using a common design for all hypotheses

# Average information fraction across hypotheses
IF_avg <- mean(c(49 / 123, 51 / 128, 50 / 126, 52 / 130))

gs_design <- gsDesign(
  k = 2,
  test.type = 4, # Non-binding futility
  alpha = 0.025 / 4, # Per-hypothesis alpha (Bonferroni for single hypothesis)
  beta = 0.1,
  timing = IF_avg,
  sfu = sfLDOF,   # Lan-DeMets O'Brien-Fleming for efficacy
  sfl = sfHSD,     # HSD for futility
  sflpar = -2
)

# Futility bound (on p-value scale)
futility_z <- gs_design$lower$bound[1]
futility_p <- pnorm(futility_z)

cat("Information fraction:", round(IF_avg, 3), "\n")
cat("Futility Z-bound at IA:", round(futility_z, 4), "\n")
cat("Futility p-value bound at IA:", round(futility_p, 4), "\n")
cat("Interpretation: Stop for futility if one-sided p-value >",
    round(futility_p, 4), "\n")
```

```{r}
# Display the gsDesign summary
tibble(
  Analysis = c("IA", "FA"),
  `Information Fraction` = round(c(IF_avg, 1), 3),
  `Futility Z-bound` = round(gs_design$lower$bound, 4),
  `Futility p-value` = round(pnorm(gs_design$lower$bound), 4)
) %>%
  gt() %>%
  tab_header(title = "Non-binding Futility Bounds (Common Across All Doses)")
```

The futility bound is applied as a common bar: at the interim analysis,
any dose arm with a one-sided p-value exceeding the futility threshold
is dropped for futility. This is a non-binding bound, meaning it does
not affect the Type I error control of the efficacy bounds.

# Efficacy Bounds

## Information Fractions

Each hypothesis has its own information fraction based on the ratio of
events at the interim to final analysis.

```{r}
# Information fractions for each hypothesis
IF_H1 <- c(49 / 123, 1)
IF_H2 <- c(51 / 128, 1)
IF_H3 <- c(50 / 126, 1)
IF_H4 <- c(52 / 130, 1)

tibble(
  Hypothesis = paste0("H", 1:4),
  `IF at IA` = round(c(IF_H1[1], IF_H2[1], IF_H3[1], IF_H4[1]), 3),
  `IF at FA` = 1
) %>%
  gt() %>%
  tab_header(title = "Information Fractions")
```

## Bonferroni Bounds

We first compute Bonferroni bounds (Type 0) as a reference, using
Lan-DeMets O'Brien-Fleming spending for each hypothesis.

```{r}
# Bonferroni bounds
bound_Bonf <- generate_bounds(
  type = 0, k = 2, w = w, m = m,
  corr = corr, alpha = 0.025,
  sf = list(sfLDOF, sfLDOF, sfLDOF, sfLDOF),
  sfparm = list(0, 0, 0, 0),
  t = list(IF_H1, IF_H2, IF_H3, IF_H4)
)

bound_Bonf %>%
  gt() %>%
  fmt_number(columns = 3:6, decimals = 6) %>%
  tab_header(title = "Bonferroni Bounds (Type 0)")
```

## WPGSD Bounds

We compute WPGSD bounds using $\alpha$-spending approach 2 (Type 3,
separate spending per hypothesis), which applies an inflation factor
$\xi$ to the Bonferroni nominal boundaries.

```{r}
set.seed(1234)

# WPGSD bounds with separate alpha spending (Type 3)
bound_WPGSD <- generate_bounds(
  type = 3, k = 2, w = w, m = m,
  corr = corr, alpha = 0.025,
  sf = list(sfLDOF, sfLDOF, sfLDOF, sfLDOF),
  sfparm = list(0, 0, 0, 0),
  t = list(IF_H1, IF_H2, IF_H3, IF_H4)
)

bound_WPGSD %>%
  gt() %>%
  fmt_number(columns = 3:6, decimals = 6) %>%
  tab_header(title = "WPGSD Bounds (Type 3, Separate Spending)")
```

## Comparison: Bonferroni vs. WPGSD

```{r, echo=FALSE}
# Combine bounds for comparison
bounds_compare <- left_join(bound_Bonf, bound_WPGSD,
  by = c("Hypotheses", "Analysis"),
  suffix = c(".B", ".W")
)

bounds_compare %>%
  gt() %>%
  fmt_number(columns = where(is.numeric), decimals = 6) %>%
  tab_header(title = "Bonferroni vs. WPGSD Bounds")
```

The WPGSD bounds are less conservative than the Bonferroni bounds
because they account for the positive correlation induced by the shared
control arm. With 4 dose groups sharing a common control, the
correlations are substantial (~0.50 within the same analysis), and the
WPGSD approach provides a meaningful improvement.

# Combined Efficacy and Futility Summary

```{r}
# Extract elementary hypothesis bounds at each analysis for both methods
elementary_hyps <- c("H1", "H2", "H3", "H4")

# Get WPGSD bounds for elementary hypotheses
wpgsd_elem <- bound_WPGSD %>%
  filter(Hypotheses %in% elementary_hyps)

bonf_elem <- bound_Bonf %>%
  filter(Hypotheses %in% elementary_hyps)

# Summary table
summary_tbl <- tibble(
  Hypothesis = rep(elementary_hyps, each = 2),
  Analysis = rep(c("IA", "FA"), 4),
  `Bonferroni p-bound` = NA_real_,
  `WPGSD p-bound` = NA_real_,
  `Futility p-bound` = NA_real_
)

for (i in seq_along(elementary_hyps)) {
  h <- elementary_hyps[i]
  bonf_row <- bonf_elem %>% filter(Hypotheses == h)
  wpgsd_row <- wpgsd_elem %>% filter(Hypotheses == h)

  for (k in 1:2) {
    row_idx <- (i - 1) * 2 + k
    summary_tbl$`Bonferroni p-bound`[row_idx] <- as.numeric(bonf_row[bonf_row$Analysis == k, h])
    summary_tbl$`WPGSD p-bound`[row_idx] <- as.numeric(wpgsd_row[wpgsd_row$Analysis == k, h])
    if (k == 1) {
      summary_tbl$`Futility p-bound`[row_idx] <- futility_p
    } else {
      summary_tbl$`Futility p-bound`[row_idx] <- NA
    }
  }
}

summary_tbl %>%
  gt() %>%
  fmt_number(columns = 3:5, decimals = 6) %>%
  sub_missing(missing_text = "-") %>%
  tab_header(
    title = "Combined Efficacy and Futility Bounds",
    subtitle = "Futility applied as a common non-binding bar at IA"
  )
```

# Closed Testing

We demonstrate the closed testing procedure with hypothetical observed
p-values. Suppose at the interim analysis we observe promising results
for Doses 2 and 4 but not for Doses 1 and 3. At the final analysis,
Dose 2 confirms its efficacy.

```{r}
# Observed p-values (hypothetical)
p_obs <- tribble(
  ~Analysis, ~H1, ~H2, ~H3, ~H4,
  1, 0.20, 0.003, 0.15, 0.005,
  2, 0.10, 0.001, 0.08, 0.003
)

p_obs %>%
  gt() %>%
  fmt_number(columns = 2:5, decimals = 6, drop_trailing_zeros = TRUE) %>%
  tab_header(title = "Observed Nominal p-Values")
```

```{r}
# Apply closed testing procedure
test_result <- closed_test(bound_WPGSD, p_obs)

test_result %>%
  gt() %>%
  tab_header(title = "Closed Testing Results (WPGSD)")
```

```{r}
# Compare with Bonferroni closed testing
test_result_bonf <- closed_test(bound_Bonf, p_obs)

test_result_bonf %>%
  gt() %>%
  tab_header(title = "Closed Testing Results (Bonferroni)")
```

## Futility Assessment

At the interim analysis, we also check each dose against the common
futility bar:

```{r}
# Futility assessment at IA
futility_check <- tibble(
  Hypothesis = paste0("H", 1:4),
  `Observed p (IA)` = c(0.20, 0.003, 0.15, 0.005),
  `Futility bound` = futility_p,
  `Drop for futility?` = ifelse(
    c(0.20, 0.003, 0.15, 0.005) > futility_p,
    "Yes", "No"
  )
)

futility_check %>%
  gt() %>%
  fmt_number(columns = 2:3, decimals = 4) %>%
  tab_header(
    title = "Futility Assessment at Interim Analysis",
    subtitle = paste("Common futility p-value threshold:", round(futility_p, 4))
  )
```

# Discussion

This vignette demonstrates the WPGSD approach for a Dunnett-type
dose-finding trial with 4 experimental arms versus a common control.
Key observations:

1.  **Correlation structure**: The shared control arm induces positive
    correlations of approximately 0.50 between test statistics for
    different dose-versus-control comparisons at the same analysis. This
    correlation is substantial and directly exploitable by the WPGSD
    approach.

2.  **Bound improvement**: The WPGSD bounds are less conservative than
    Bonferroni bounds, providing increased nominal significance levels
    at both interim and final analyses. This gain comes at no cost to
    FWER control.

3.  **Scalability**: The event table and correlation formula scale
    naturally to any number of dose groups. With $M$ hypotheses, the
    event table requires $M(M+1)/2$ entries per analysis (4 diagonal + 6
    off-diagonal for $M = 4$).

4.  **Futility**: A common non-binding futility bound is computed
    externally via `gsDesign` and applied as a screening criterion at
    the interim analysis. Since the futility bound is non-binding, it
    does not affect the WPGSD efficacy bounds or FWER control.

5.  **Closed testing**: The closed testing procedure is necessary for
    FWER control because consonance does not always hold for the WPGSD
    approach. With 4 hypotheses, there are $2^4 - 1 = 15$ intersection
    hypotheses to evaluate.

# References
