---
title: "Dunnett-type dose-finding with group sequential design"
author: "Keaven Anderson"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
bibliography: wpgsd.bib
vignette: >
  %\VignetteIndexEntry{Dunnett-type dose-finding with group sequential design}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  fig.align = "center"
)
```

```{r, message=FALSE, warning=FALSE}
library(wpgsd)
library(gsDesign)
library(gMCPLite)
library(dplyr)
library(tibble)
library(gt)
```

# Background

A common clinical trial design compares multiple experimental dose
groups against a shared control arm. This is known as a Dunnett-type
design (@dunnett1955multiple). When combined with a group sequential
design allowing interim analyses, the correlation between test
statistics for different dose-versus-control comparisons arises
naturally from the shared control arm.

The weighted parametric group sequential design (WPGSD) approach
(@anderson2022unified) exploits these known correlations to produce
tighter efficacy bounds compared to the Bonferroni approach. This
vignette demonstrates the WPGSD methodology for a trial with 2
experimental dose groups versus a common control, using overall survival
(OS) as the primary endpoint, with a group sequential design having two
interim analyses and a final analysis ($K = 3$). Using $K > 2$ analyses
exercises the full generality of the correlation computation formula.

# Trial Setup

## Hypotheses

We consider 2 hypotheses, each comparing an experimental dose group to
the common control:

-   $H_1$: Dose 1 vs. Control
-   $H_2$: Dose 2 vs. Control

## Multiplicity Strategy

We use equal initial weights ($w_i = 1/2$ for each hypothesis) and
full transitions ($m_{12} = m_{21} = 1$). This symmetric strategy
reflects no prior preference between doses.

```{r, out.width="90%"}
# Transition matrix: full transition between 2 hypotheses
m <- matrix(c(0, 1, 1, 0), nrow = 2, byrow = TRUE)

# Initial weights: equal allocation
w <- c(1 / 2, 1 / 2)

# Multiplicity graph
cbPalette <- c("#E69F00", "#56B4E9")

nameHypotheses <- c(
  "H1: Dose 1 vs Control",
  "H2: Dose 2 vs Control"
)

hplot <- hGraph(2,
  alphaHypotheses = w,
  m = m,
  nameHypotheses = nameHypotheses,
  trhw = .2, trhh = .1,
  digits = 4, trdigits = 4, size = 5, halfWid = 1.2,
  halfHgt = 0.5, offset = 0.2, trprop = 0.35,
  fill = as.factor(1:2),
  palette = cbPalette,
  wchar = "w"
)
hplot
```

## Event Counts

We assume approximately balanced event counts across treatment arms
with three analyses: two interim analyses (IA1 at ~30% information,
IA2 at ~60% information) and a final analysis (FA). For OS, events
are deaths observed in each dose-versus-control comparison.

| Treatment Arm | Events at IA1 | Events at IA2 | Events at FA |
|:-------------:|:-------------:|:-------------:|:------------:|
| Dose 1        | 20            | 40            | 65           |
| Dose 2        | 22            | 44            | 70           |
| Control       | 21            | 42            | 67           |

Table: Number of events at each analysis for each treatment arm.
IA1: first interim (~30% information). IA2: second interim (~60%
information). FA: final analysis.

The diagonal entries of the event table represent total events for each
hypothesis (experimental arm + control), while off-diagonal entries
represent events shared between two comparisons (control arm events
only).

```{r}
# Event counts per arm
# IA1: Dose 1=20, Dose 2=22, Control=21
# IA2: Dose 1=40, Dose 2=44, Control=42
# FA:  Dose 1=65, Dose 2=70, Control=67

# Event table: diagonal = experimental + control; off-diagonal = control only
event <- tribble(
  ~H1, ~H2, ~Analysis, ~Event,
  # IA1 diagonal (total events per hypothesis = dose + control)
  1, 1, 1, 41, # Dose 1 (20) + Control (21)
  2, 2, 1, 43, # Dose 2 (22) + Control (21)
  # IA1 off-diagonal (shared events = control only)
  1, 2, 1, 21,
  # IA2 diagonal
  1, 1, 2, 82, # Dose 1 (40) + Control (42)
  2, 2, 2, 86, # Dose 2 (44) + Control (42)
  # IA2 off-diagonal
  1, 2, 2, 42,
  # FA diagonal
  1, 1, 3, 132, # Dose 1 (65) + Control (67)
  2, 2, 3, 137, # Dose 2 (70) + Control (67)
  # FA off-diagonal
  1, 2, 3, 67
)

event %>%
  gt() %>%
  tab_header(title = "Event Count Table")
```

# Correlation Structure

## Computing Correlations

The correlation between test statistics follows the formula
(@anderson2022unified):

$$\text{Corr}(Z_{ik}, Z_{i'k'}) = \frac{n_{i \wedge i', k \wedge k'}}{\sqrt{n_{ik} \cdot n_{i'k'}}}$$

where $n_{i \wedge i', k \wedge k'}$ denotes the number of events
included in both $Z_{ik}$ and $Z_{i'k'}$.

For the Dunnett-type design, the shared events between two different
dose-versus-control comparisons at the same analysis are the control
arm events. For example, at the first interim analysis:

$$\text{Corr}(Z_{1,1}, Z_{2,1}) = \frac{n_{\text{control, IA1}}}{\sqrt{n_{1,1} \cdot n_{2,1}}} = \frac{21}{\sqrt{41 \cdot 43}} \approx 0.500$$

With three analyses, the correlation matrix is $6 \times 6$ (2
hypotheses $\times$ 3 analyses). This exercises the general
correlation formula for all four cases:

1.  **Same hypothesis, same analysis**: $\text{Corr} = 1$
2.  **Same hypothesis, different analyses** ($k < k'$):
    $\text{Corr}(Z_{i,k}, Z_{i,k'}) = \sqrt{n_{ik}/n_{ik'}}$
3.  **Different hypotheses, same analysis**:
    $\text{Corr}(Z_{i,k}, Z_{i',k}) = n_{\text{control},k}/\sqrt{n_{ik} \cdot n_{i'k}}$
4.  **Different hypotheses, different analyses** ($k < k'$):
    $\text{Corr}(Z_{i,k}, Z_{i',k'}) = n_{\text{control},k}/\sqrt{n_{ik} \cdot n_{i'k'}}$

```{r}
# Generate correlation matrix
corr <- generate_corr(event)

corr %>%
  as_tibble() %>%
  gt() %>%
  fmt_number(columns = everything(), decimals = 3) %>%
  tab_header(title = "Correlation Matrix (6 x 6)")
```

## Verification of Correlation Entries

We manually verify key entries of the correlation matrix to confirm
correctness for the 2-hypothesis, 3-analysis scenario.

```{r}
# --- Verification of key correlation entries ---

# Case 1: Same hypothesis, same analysis -> should be 1.0
cat("Corr(Z_1,1, Z_1,1) =", corr[1, 1], " (expected: 1.0)\n")

# Case 2: Different hypotheses, same analysis (IA1)
# Corr(Z_1,1, Z_2,1) = n_control_IA1 / sqrt(n_H1_IA1 * n_H2_IA1) = 21 / sqrt(41 * 43)
expected_12_ia1 <- 21 / sqrt(41 * 43)
cat(
  "Corr(Z_1,1, Z_2,1) =", round(corr[1, 2], 6),
  " (expected:", round(expected_12_ia1, 6), ")\n"
)

# Case 3: Different hypotheses, same analysis (FA)
# Corr(Z_1,3, Z_2,3) = n_control_FA / sqrt(n_H1_FA * n_H2_FA) = 67 / sqrt(132 * 137)
expected_12_fa <- 67 / sqrt(132 * 137)
cat(
  "Corr(Z_1,3, Z_2,3) =", round(corr[5, 6], 6),
  " (expected:", round(expected_12_fa, 6), ")\n"
)

# Case 4: Same hypothesis, different analyses (IA1 vs IA2)
# Corr(Z_1,1, Z_1,2) = sqrt(n_H1_IA1 / n_H1_IA2) = sqrt(41 / 82)
expected_11_ia1_ia2 <- sqrt(41 / 82)
cat(
  "Corr(Z_1,1, Z_1,2) =", round(corr[1, 3], 6),
  " (expected:", round(expected_11_ia1_ia2, 6), ")\n"
)

# Case 5: Same hypothesis, different analyses (IA1 vs FA)
# Corr(Z_1,1, Z_1,3) = sqrt(n_H1_IA1 / n_H1_FA) = sqrt(41 / 132)
expected_11_ia1_fa <- sqrt(41 / 132)
cat(
  "Corr(Z_1,1, Z_1,3) =", round(corr[1, 5], 6),
  " (expected:", round(expected_11_ia1_fa, 6), ")\n"
)

# Case 6: Same hypothesis, different analyses (IA2 vs FA)
# Corr(Z_1,2, Z_1,3) = sqrt(n_H1_IA2 / n_H1_FA) = sqrt(82 / 132)
expected_11_ia2_fa <- sqrt(82 / 132)
cat(
  "Corr(Z_1,2, Z_1,3) =", round(corr[3, 5], 6),
  " (expected:", round(expected_11_ia2_fa, 6), ")\n"
)

# Case 7: Different hypotheses, different analyses (IA1 vs IA2)
# Corr(Z_1,1, Z_2,2) = n_control_IA1 / sqrt(n_H1_IA1 * n_H2_IA2) = 21 / sqrt(41 * 86)
expected_12_ia1_ia2 <- 21 / sqrt(41 * 86)
cat(
  "Corr(Z_1,1, Z_2,2) =", round(corr[1, 4], 6),
  " (expected:", round(expected_12_ia1_ia2, 6), ")\n"
)

# Case 8: Different hypotheses, different analyses (IA2 vs FA)
# Corr(Z_1,2, Z_2,3) = n_control_IA2 / sqrt(n_H1_IA2 * n_H2_FA) = 42 / sqrt(82 * 137)
expected_12_ia2_fa <- 42 / sqrt(82 * 137)
cat(
  "Corr(Z_1,2, Z_2,3) =", round(corr[3, 6], 6),
  " (expected:", round(expected_12_ia2_fa, 6), ")\n"
)

# Case 9: Different hypotheses, different analyses (IA1 vs FA)
# Corr(Z_2,1, Z_1,3) = n_control_IA1 / sqrt(n_H2_IA1 * n_H1_FA) = 21 / sqrt(43 * 132)
expected_21_ia1_fa <- 21 / sqrt(43 * 132)
cat(
  "Corr(Z_2,1, Z_1,3) =", round(corr[2, 5], 6),
  " (expected:", round(expected_21_ia1_fa, 6), ")\n"
)
```

```{r}
# Verify matrix is positive definite
eigenvalues <- eigen(corr)$values
cat("Eigenvalues:", round(eigenvalues, 6), "\n")
cat("All positive:", all(eigenvalues > 0), "\n")
```

The correlation matrix for this 2-hypothesis, 3-analysis Dunnett
scenario has the following structure:

-   **Same analysis, different hypotheses**: Correlation $\approx 0.50$,
    driven by the shared control arm. With balanced arms, this
    approaches $n_C / (n_C + n_E) \approx 0.5$ when $n_C \approx n_E$.
-   **Same hypothesis, adjacent analyses**: Correlation
    $\approx 0.71$ (IA1 vs IA2) and $\approx 0.79$ (IA2 vs FA),
    following $\sqrt{n_{ik}/n_{ik'}}$.
-   **Same hypothesis, non-adjacent analyses**: Correlation
    $\approx 0.56$ (IA1 vs FA), demonstrating the general formula for
    $K > 2$ analyses.
-   **Different hypotheses, different analyses**: Correlation
    $\approx 0.35$ (IA1 vs IA2) and $\approx 0.40$ (IA2 vs FA),
    combining both sources of correlation.

# Futility Bounds

We use `gsDesign` to compute common non-binding futility bounds at the
interim analyses. These bounds apply equally to both dose arms. The
futility bounds are based on a beta-spending approach using a
Hwang-Shih-DeCani spending function.

```{r}
# Group sequential design with non-binding futility
# Average information fraction across hypotheses at each analysis
IF_avg <- c(
  mean(c(41 / 132, 43 / 137)),
  mean(c(82 / 132, 86 / 137)),
  1
)

gs_design <- gsDesign(
  k = 3,
  test.type = 4, # Non-binding futility
  alpha = 0.025 / 2, # Per-hypothesis alpha (Bonferroni for single hypothesis)
  beta = 0.1,
  timing = IF_avg[1:2],
  sfu = sfLDOF, # Lan-DeMets O'Brien-Fleming for efficacy
  sfl = sfHSD, # HSD for futility
  sflpar = -2
)

# Futility bounds (on p-value scale)
futility_z <- gs_design$lower$bound
futility_p <- pnorm(futility_z)

cat("Information fractions:", round(IF_avg, 3), "\n")
cat("Futility Z-bounds:", round(futility_z, 4), "\n")
cat("Futility p-value bounds:", round(futility_p, 4), "\n")
```

```{r}
# Display the gsDesign summary
tibble(
  Analysis = c("IA1", "IA2", "FA"),
  `Information Fraction` = round(IF_avg, 3),
  `Futility Z-bound` = round(gs_design$lower$bound, 4),
  `Futility p-value` = round(pnorm(gs_design$lower$bound), 4)
) %>%
  gt() %>%
  tab_header(title = "Non-binding Futility Bounds (Common Across Both Doses)")
```

The futility bound is applied as a common bar: at each interim
analysis, any dose arm with a one-sided p-value exceeding the futility
threshold may be dropped for futility. This is a non-binding bound,
meaning it does not affect the Type I error control of the efficacy
bounds.

# Efficacy Bounds

## Information Fractions

Each hypothesis has its own information fraction based on the ratio of
events at each analysis to the final analysis.

```{r}
# Information fractions for each hypothesis
IF_H1 <- c(41 / 132, 82 / 132, 1)
IF_H2 <- c(43 / 137, 86 / 137, 1)

tibble(
  Hypothesis = c("H1", "H2"),
  `IF at IA1` = round(c(IF_H1[1], IF_H2[1]), 3),
  `IF at IA2` = round(c(IF_H1[2], IF_H2[2]), 3),
  `IF at FA` = 1
) %>%
  gt() %>%
  tab_header(title = "Information Fractions")
```

## Bonferroni Bounds

We first compute Bonferroni bounds (Type 0) as a reference, using
Lan-DeMets O'Brien-Fleming spending for each hypothesis.

```{r}
# Bonferroni bounds
bound_Bonf <- generate_bounds(
  type = 0, k = 3, w = w, m = m,
  corr = corr, alpha = 0.025,
  sf = list(sfLDOF, sfLDOF),
  sfparm = list(0, 0),
  t = list(IF_H1, IF_H2)
)

bound_Bonf %>%
  gt() %>%
  fmt_number(columns = 3:4, decimals = 6) %>%
  tab_header(title = "Bonferroni Bounds (Type 0)")
```

## WPGSD Bounds

We compute WPGSD bounds using $\alpha$-spending approach 2 (Type 3,
separate spending per hypothesis), which applies an inflation factor
$\xi$ to the Bonferroni nominal boundaries.

```{r}
set.seed(1234)

# WPGSD bounds with separate alpha spending (Type 3)
bound_WPGSD <- generate_bounds(
  type = 3, k = 3, w = w, m = m,
  corr = corr, alpha = 0.025,
  sf = list(sfLDOF, sfLDOF),
  sfparm = list(0, 0),
  t = list(IF_H1, IF_H2)
)

bound_WPGSD %>%
  gt() %>%
  fmt_number(columns = 3:4, decimals = 6) %>%
  tab_header(title = "WPGSD Bounds (Type 3, Separate Spending)")
```

## Comparison: Bonferroni vs. WPGSD

```{r, echo=FALSE}
# Combine bounds for comparison
bounds_compare <- left_join(bound_Bonf, bound_WPGSD,
  by = c("Hypotheses", "Analysis"),
  suffix = c(".B", ".W")
)

bounds_compare %>%
  gt() %>%
  fmt_number(columns = where(is.numeric), decimals = 6) %>%
  tab_header(title = "Bonferroni vs. WPGSD Bounds")
```

The WPGSD bounds are less conservative than the Bonferroni bounds
because they account for the positive correlation induced by the shared
control arm. With 2 dose groups sharing a common control, the
correlations within the same analysis are substantial (~0.50), and the
WPGSD approach provides a meaningful improvement.

# Combined Efficacy and Futility Summary

```{r}
# Extract elementary hypothesis bounds at each analysis for both methods
elementary_hyps <- c("H1", "H2")

# Get WPGSD bounds for elementary hypotheses
wpgsd_elem <- bound_WPGSD %>%
  filter(Hypotheses %in% elementary_hyps)

bonf_elem <- bound_Bonf %>%
  filter(Hypotheses %in% elementary_hyps)

# Summary table
summary_tbl <- tibble(
  Hypothesis = rep(elementary_hyps, each = 3),
  Analysis = rep(c("IA1", "IA2", "FA"), 2),
  `Bonferroni p-bound` = NA_real_,
  `WPGSD p-bound` = NA_real_,
  `Futility p-bound` = NA_real_
)

for (i in seq_along(elementary_hyps)) {
  h <- elementary_hyps[i]
  bonf_row <- bonf_elem %>% filter(Hypotheses == h)
  wpgsd_row <- wpgsd_elem %>% filter(Hypotheses == h)

  for (k in 1:3) {
    row_idx <- (i - 1) * 3 + k
    summary_tbl$`Bonferroni p-bound`[row_idx] <- as.numeric(bonf_row[bonf_row$Analysis == k, h])
    summary_tbl$`WPGSD p-bound`[row_idx] <- as.numeric(wpgsd_row[wpgsd_row$Analysis == k, h])
    if (k <= 2) {
      summary_tbl$`Futility p-bound`[row_idx] <- futility_p[k]
    } else {
      summary_tbl$`Futility p-bound`[row_idx] <- NA
    }
  }
}

summary_tbl %>%
  gt() %>%
  fmt_number(columns = 3:5, decimals = 6) %>%
  sub_missing(missing_text = "-") %>%
  tab_header(
    title = "Combined Efficacy and Futility Bounds",
    subtitle = "Futility applied as common non-binding bars at interim analyses"
  )
```

# Closed Testing

We demonstrate the closed testing procedure with hypothetical observed
p-values. Suppose at IA1 we see a strong signal for Dose 2 but not
Dose 1. At IA2 both doses show improvement. At the final analysis,
Dose 2 confirms strong efficacy.

```{r}
# Observed p-values (hypothetical)
p_obs <- tribble(
  ~Analysis, ~H1, ~H2,
  1, 0.20, 0.004,
  2, 0.05, 0.002,
  3, 0.02, 0.001
)

p_obs %>%
  gt() %>%
  fmt_number(columns = 2:3, decimals = 6, drop_trailing_zeros = TRUE) %>%
  tab_header(title = "Observed Nominal p-Values")
```

```{r}
# Apply closed testing procedure
test_result <- closed_test(bound_WPGSD, p_obs)

test_result %>%
  gt() %>%
  tab_header(title = "Closed Testing Results (WPGSD)")
```

```{r}
# Compare with Bonferroni closed testing
test_result_bonf <- closed_test(bound_Bonf, p_obs)

test_result_bonf %>%
  gt() %>%
  tab_header(title = "Closed Testing Results (Bonferroni)")
```

## Futility Assessment

At each interim analysis, we also check each dose against the common
futility bar:

```{r}
# Futility assessment at IA1 and IA2
futility_check <- tibble(
  Hypothesis = rep(c("H1", "H2"), 2),
  Analysis = rep(c("IA1", "IA2"), each = 2),
  `Observed p` = c(0.20, 0.004, 0.05, 0.002),
  `Futility bound` = rep(futility_p[1:2], each = 2),
  `Drop for futility?` = ifelse(
    c(0.20, 0.004, 0.05, 0.002) > rep(futility_p[1:2], each = 2),
    "Yes", "No"
  )
)

futility_check %>%
  gt() %>%
  fmt_number(columns = 3:4, decimals = 4) %>%
  tab_header(
    title = "Futility Assessment at Interim Analyses",
    subtitle = "Common non-binding futility threshold at each interim"
  )
```

# Discussion

This vignette demonstrates the WPGSD approach for a Dunnett-type
dose-finding trial with 2 experimental arms versus a common control
and 3 analyses (2 interims + final). Key observations:

1.  **Correlation structure**: The shared control arm induces positive
    correlations of approximately 0.50 between test statistics for
    different dose-versus-control comparisons at the same analysis.
    This correlation is substantial and directly exploitable by the
    WPGSD approach.

2.  **Three analyses**: Using $K = 3$ analyses demonstrates the full
    generality of the correlation formula. The within-hypothesis
    correlations across non-adjacent analyses (e.g., IA1 vs. FA)
    are correctly computed as $\sqrt{n_{ik}/n_{ik'}}$, and the
    cross-hypothesis, cross-analysis correlations use the shared
    events at the earlier analysis.

3.  **Bound improvement**: The WPGSD bounds are less conservative than
    Bonferroni bounds, providing increased nominal significance levels
    at all three analyses. This gain comes at no cost to FWER control.

4.  **Futility**: Common non-binding futility bounds are computed
    externally via `gsDesign` with $K = 3$ and applied as screening
    criteria at each interim analysis. Since the futility bounds are
    non-binding, they do not affect the WPGSD efficacy bounds or FWER
    control.

5.  **Closed testing**: The closed testing procedure is necessary for
    FWER control. With 2 hypotheses, there are $2^2 - 1 = 3$
    intersection hypotheses to evaluate, making the procedure
    straightforward.

# References
